---
title: "'Hot Rocks That Shoot Ghost Bullets': Native American Perceptions of a Nuclear Waste Facility"
abstract: Environmental justice concerns are raised when environmental risks and hazards are inequitably distributed across society. Native American populations have long been at the center of environmental justice disputes, however relatively little quantitative research has examined how Native Americans view the risks related to the siting of a potentially hazardous facility. Combining 35 statewide surveys collected from 1990 to 2001, the following study explores the risk perceptions of Native American populations in New Mexico regarding the storage of nuclear waste. We find that Native Americans tend to be less supportive of the Waste Isolation Pilot Plant (WIPP) facility, and have higher perceptions of risk than white respondents. We also explore intraracial variation within Native American populations and find that older males tended to be more supportive of WIPP, and perceive less risk. These findings help in understanding how Native Americans perceive risk as it relates to environmental issues.
...

```{r, include=FALSE}
wipp <- read.csv("wippDAT.csv")
``` 

# Introduction

Native American populations have a particularly long and complicated history with issues of environmental justice and degradation within their communities. This history is likely to weigh heavily on their collective perception of risk related to potential environmental hazards. Also, Native Americans are members of distinct Native nations, which places these communities in a unique position relative to other actors in the intergovernmental relations framework concerning environmental risk management [@wilkins_american_2010]. As a result, when state and local governments have been opposed to federal programs such as those related to nuclear waste storage within their communities, tribal governments have been seen as another viable contender for federal funds to locate facilities deemed undesirable by other jurisdictions [@gedicks_new_1994; @gowda_voluntary_2000]. Such targeting has naturally raised concerns for environmental justice and equity as Native communities who, in some instances, face extreme levels of poverty and other social maladies and, as a result, may be more likely to accept such high-risk programs in light of potential economic returns [@lynch_native_2012]. As @bullard_environmentalism_2000 [pg.571] argues, "Native American nations have become prime targets for waste trading." These dynamics make it even more important to understand how Native Americans perceive levels of risk that have thus far gone relatively unexplored in the literature on environmental justice. 

While the environmental risk perceptions of Native Americans may not be well understood, differences in risk perceptions across other demographic groups have been demonstrated across a wide range of issues. One of the major insights offered by this research is that white–males tend to hold lower perceptions of risk. The "white-male effect" is particularly pronounced with respect to environmental hazards, where white males have consistently been found to perceive environmental and/or technological risks as smaller and more acceptable than other demographic groups [@flynn_gender_1994; @davidson_gender_1996; @bord_gender_1997; @gustafson_gender_1998; @slovic_trust_1999; @finucane_gender_2000; @kalof_race_2002; @satterfield_discrimination_2004]. In general, white–males with higher levels of risk acceptance tend to be better educated, have higher incomes, and are more politically conservative than other groups [@flynn_gender_1994; @mccright_bringing_2013]. In addition, white–males also tend to place a higher degree of trust in institutions and authority to effectively manage risk [@flynn_gender_1994]. 

Much of the work comparing the risk perceptions of white-males to minority groups has focused on African–American, Hispanic, and Asian populations. In addition, this work often assumes homogeneity among minority groups, with little consideration given to differences across individuals within these groups. The assumption of homogeneity may mask important differences within ethic groups in their perceptions of risk. For example, @rivers_simple_2010 found that African-American males had lower levels of perceived risks than African-American females, thus demonstrating the "white-male" effect within another demographic group. Given the history of potential exposure to hazards experienced by Native American populations and the subsequent environmental justice concerns [@gowda_voluntary_2000], work focused on environmental or technological risk perceptions of Native Americans is needed. Using a combination of public opinion surveys conducted in New Mexico from 1990 to 2001, this paper examines the risk perceptions of Native Americans related to the Waste Isolation Pilot Plant (WIPP), a low-level radioactive waste facility located in southeast New Mexico. Specifically, we 1) compare the views of Native Americans about WIPP to other demographic groups and 2) we explore differences among Native Americans concerning WIPP. 

After almost a quarter century of debate, the WIPP facility in New Mexico opened and began receiving shipments of transuranic defense-related waste in the spring of 1999. Prior to its opening, the WIPP facility was highly controversial and contested by many state officials and the public [@mccutcheon_nuclear_2002]. However, opposition to the facility from New Mexico residents steadily declined over time and by 2001 the facility enjoyed support from a majority of residents [@jenkins-smith_reversing_2011]. Given that New Mexico has a sizable native population, the WIPP facility is an ideal case for measuring risk perceptions within that population. Indeed, the native populations of New Mexico have a troubled history with nuclear technology, most notably, the Navajo nation with uranium mining [@brugge_history_2002] and the Mescalero Apaches with the Monitored-Retrievable Storage (MRS) of nuclear waste [@gowda_voluntary_2000]. The Mescalero Apaches did not have a term in their language that represented radiation, so during the discussions of MRS storage they termed nuclear waste as "hot rocks that shoot ghost bullets" [@adams_biologist_2014]. Finally, the large-_N_ nature of the data available for this study allows for one of the first quantitative comparisons between Native Americans and other ethnic groups on energy and environmental policy, as well _intraracial_ comparisons between self-identified Native Americans. 


# Native Americans, Environmental Justice, and Risk Perceptions

Perceptions of the risks associated with nuclear energy and nuclear waste storage vary considerably across the population. Policy debates surrounding nuclear energy development have evoked concerns regarding general public safety and inter-generational equity and have greatly shaped the way the public thinks about the potential hazards associated with nuclear energy [@ansolabehere_public_2009]. In addition, nuclear energy development raises concerns regarding environmental justice, with the potential for underrepresented communities to be exposed to disproportionate risks. Native nations, particularly in the American southwest, have been at the center of federal nuclear energy programs, including uranium mining and nuclear waste storage [@reed_native_2002]. The role of native communities in the development of nuclear energy raises environmental justice concerns along several dimensions, and each of these dimensions could impact the ways in which Native Americans view a potential nuclear waste facility. Despite these concerns little large-_N_ research has examined the views of the native population regarding environmental risks or nuclear issues more specifically.

Scholars have noted several factors that influence the ways in which the general public view nuclear energy issues and these include demographics, proximity to the facility, values, and trust in government. Each of these correlate with higher or lower perceived levels of risk and help to explain support or opposition to nuclear programs. While previous quantitative research has not examined Native Americans specifically, the experiences of tribal communities with nuclear energy and waste raise, at least, three environmental justice concerns that could offer insights into how Native Americans may view a potential nuclear waste facility. These concerns include equity and representation; historical environmental and health hazards on tribal lands; and distrust of the American federal government. The risk perception literature, combined with notions of environmental justice, can be used to generate testable hypotheses regarding Native American views of the WIPP facility in New Mexico. We will briefly describe some of the insights provided by the risk perception literature, coupled with a discussion of how that literature can inform the analysis of environmental justice related to Native Americans and nuclear energy. Following that discussion, we posit three hypotheses about the views of Native Americans in New Mexico regarding the Waste Isolation Pilot Plant (WIPP), a facility designed for the long-term storage of defense-related nuclear waste. 

The first environmental justice concern regards the equitable distribution of risk across society. Environmental justice is broadly understood to be defined as a concern about the unequal distribution of risk, where poor and minority communities are more likely than other demographic groups to be exposed to environmental risks such as air pollution [@clark_national_2014], and proximity to Superfund sites [@stretesky_environmental_1998]. Related to this dimension of environmental justice is the consistent finding that demographics such as age, income, gender, education level, and race play a large role in determining perceptions of risk [@finucane_gender_2000]. In particular, a rich body of research has focused on the presence of a white-male effect with regard to risk perception across a wide variety of potential hazards, where white-males tend to perceive lower levels of risk and are more supportive of potentially hazardous facilities than women or racial minorities [@flynn_gender_1994; @davidson_gender_1996; @bord_gender_1997; @gustafson_gender_1998; @slovic_trust_1999; @finucane_gender_2000; @kalof_race_2002; @satterfield_discrimination_2004; @greenberg_energy_2011; @whittaker_racialethnic_2005; @steger_gender_1989]. 

Several studies associate the white-male effect to a number of underlying factors. @flynn_gender_1994 observe that white-males with low-risk perceptions tend to have higher incomes, more education, and are politically more conservative than all other groups, and that may influence levels of perceived risk. White-males are also found to have more faith and trust in the ability of authorities to manage potential hazards. The increased trust in authority points to white-males as being more hierarchical in their worldview and, therefore, more likely to trust risk management institutions. Indeed, @kahan_culture_2007 found an interactive effect between white–males and a hierarchical-individualistic worldview when predicting risk perceptions. In addition, @finucane_gender_2000  argue that important sociopolitical characteristics help explain why differences in perceived risk exist based on gender and race. The authors argue that important cultural differences exist among white-males given their position and status in society. Whereas, women and non-white men may have less influence and power over decisions concerning potentially hazardous facilities. This contributes to a lower level of trust among women and minorities who may have fewer opportunities to guide decision-making and less access to important information in which to educate themselves about the process and efforts to mitigate risk [@mccright_bringing_2013]. @olofsson_white_2011 found no gender differences in perceptions of risk in Swedish society, which is based on gender equality, thus providing further evidence for the role of power and status in risk acceptance. Finally, it should be noted that differences between white males and other demographic groups does not imply heterogeneity in terms of drivers of risk perceptions within those other groups [@rivers_simple_2010; @greenberg_energy_2011]. For example, @rivers_simple_2010 found that African-American males were likely to have lower perceptions of risk across several potential hazards than African-American females. While the white-male effect in risk perception has been repeatedly demonstrated, there have not been any explicit comparisons with Native Americans. 

In addition to demographic factors, proximity to potentially hazardous sites has been a primary focus in exploring the attitudes of the public. A NIMBY (or "Not-in-My-Backyard") effect is thought to be present and an important driver of opposition to potentially hazardous facilities [@groothuis_locating_1994; @michaud_nimbyism_2008]. According to the standard model of NIMBYism, closer geographic proximity to undesirable or hazardous facilities should be correlated with higher levels of concern and opposition among those most likely to be directly impacted. Such concerns have been documented in studies that show a relationship between proximity to facilities and levels of opposition [@benford_whose_1993] and reduced property values in those areas [@kiel_house_1995; @gawande_nuclear_2001]. However, some studies have also found that such concerns may be mitigated by the potential economic returns from which communities close to siting areas may benefit. Contrary to expectations of NIMBYism, perceptions of benefits and familiarity with nuclear technology can lead to higher levels of support from communities closest to the effected areas  [@jenkins-smith_mitigation_2001]. Indeed, concerning nuclear technology, proximity to a nuclear plant can _increase_ support for expanding nuclear energy [@vanderpligt_attitudes_1986] and, in the case of WIPP, proximity to the facility and the transportation route was associated with increased support [@jenkins-smith_reversing_2011]. 

Given their rates of extreme poverty, tribal communities, and tribal leaders in particular, may be more willing than others to accept some environmental risks in exchange for economic development, therefore NIMBYism may not be as much of a consideration for these communities. However, this trade-off raises concerns about equity and representation when making decisions about acceptable risks for a community. These concerns were illustrated in the case of the adoption of monitored retrievable storage (MRS) facilities for high-level nuclear waste. @gowda_voluntary_2000 investigated several aspects of equity concerning the cultural perspective of Native American communities and found that after state and local officials refused to enter into voluntary agreements in response to intense opposition from key constituents, tribal governments--including the Mescalero Apache in New Mexico--were among the few actors willing to accept the risks in return for the economic incentives of the program. While the process of MRS negotiations was designed to be equitable from a largely Western context, concerns were raised as to whether decisions by tribal governments to adopt high-risk programs were truly representative of the overall sentiment of the American Indian citizenry. Thus, while tribal government representatives may be more open to high-risk facilities for primarily economic reasons, such support may mask the individual sentiments of Native Americans who may be more sensitive to greater levels of risk in similar ways as other underrepresented populations. 

Apart from NIMBYism and demographics, the core values and beliefs of individuals have been demonstrated to play a role in risk perceptions in general [@michaud_nimbyism_2008; @kahan_fear_2006] and with regard to nuclear technology specifically [@kahan_cultural_2011; @greenberg_energy_2011; @jenkins-smith_ideology_1994; @rothman_elite_1987]. In particular, differences in political values such as partisanship and ideology can influence views about nuclear energy. Liberals and Democrats have been shown to be historically more opposed to nuclear energy than are conservatives and Republicans [@rothman_elite_1987; @greenberg_energy_2011]. Also, the environmental movement has long opposed nuclear energy [@duffy_nuclear_1997]; therefore concern about the environment may be associated with less support for nuclear technology. In the case of WIPP, Republicans and conservatives were consistently more supportive than independents, liberals, and Democrats and those with lower levels of environmental concern were more supportive of the facility [@jenkins-smith_reversing_2011]. 

The core values of Native Americans are also likely to be key as many Native American groups maintain a spiritual connection with the environment and this connection may factor into their overall levels of perceived risk concerning nuclear waste storage [@bullard_environmentalism_2000; @adamson_american_2001]. Along similar lines, certain American Indian groups also take into consideration issues of inter-generational equity in their belief systems, and this can be seen in a concern for what is commonly referred to as the 7th Generation. According to this logic, it is important for human beings to consider the impact of their decisions not only on the present generation, but seven generations into the future. The condition of the environment we leave for future generations should not be discounted or dismissed, which raises the perceived costs of such risky environmental decisions as burying nuclear waste. 

Despite the environmentalism of many Native Americans, there is a well-documented history of continuing environmental degradation on tribal land, and the associated health risks are a second aspect of environment justice for tribal communities. Some of the degradation and associated health risks have resulted from uranium mining, which is a necessary component of nuclear energy production. The uranium mining that has occurred on the Navajo Reservation--located in the "four corners" region where Arizona, Colorado, New Mexico and Utah meet--has caused serious environmental and health issues for tribal residents [@brugge_history_2002]. According to the Environmental Protection Agency (EPA), approximately four million tons of uranium was mined from the Navajo reservation from the 1940s to the 1980s to help fuel the nuclear arms race with leases overseen by the U.S. federal government [@epa_addressing_2015]. Such activities continue to have devastating effects. These include the contamination of wells that provide clean drinking water to the Navajo people and livestock, a rise in cancer-related deaths and other health issues connected to exposure to radiation, and contamination of land across 520 separate sites on the reservation [@diep_abandoned_2010; @minard_navajo_2014]. In 2005, the Navajo Nation instituted a reservation-wide ban on uranium mining in light of these environmental realities though efforts to reintroduce mining in the community remain. Stories such as this are not isolated to the Navajo nation and stretch across Indian country which may arguably shape the collective assessment of risk among this particular population [@brook_environmental_1998; @frohmberg_assessment_2000; @lynch_native_2012; @nussbaum_americas_2012]. 

A final element of environmental justice that makes Native American groups potentially unique from that of other populations with relation to risk perception concerns the importance of trust in government. To say that a degree of distrust exists between Native Americans and federal officials is perhaps an understatement given the history of conflict, deception, and cultural assimilation that defined tribal-federal relations for over a century [@vine_american_1983; @wilkins_american_2010]. Scholars argue that a long line of injustices suffered at the hands of the federal government have had a dramatic effect on U.S.--tribal relations including broken treaty agreements and forced relocation in the 19th century; assimilation into Western society via federally funded and operated Indian boarding schools in the early 20th century; and a struggle for basic human rights and respect for tribal sovereignty in the 1960s to 1980s. As a result, Native Americans may be particularly suspicious of federal programs that accompany any level of potential harm such as the storage and transportation of nuclear waste. @brook_environmental_1998 goes so far as to argue that such environmental injustices are a direct extension of historical incidents of genocide against Native American people that should frame potentially damaging federal Indian policies such as mining and toxic waste. In this manner, Native Americans may be more sensitive to risk based on their collective cultural and historical experience with federal programs and policies that pose even the most remote threat to the environment. 

The importance of trust in government is not just limited to Native Americans, as levels of trust in government have been demonstrated to play a role in understanding support or opposition to facilities that accompany high levels of risk across demographic groups [@slovic_risk_1991; @groothuis_locating_1994; @siegrist_salient_2000]. These findings suggest that lower levels of trust in the government’s ability to manage hazardous facilities will lead to higher levels of perceived risk from such endeavors. This, in turn, will lead to increased opposition as groups have little faith in the ability of government officials to safely and effectively manage potentially dangerous materials. As a result, trust in government plays a pivotal role in understanding how different populations will perceive such highly controversial operations as the storage of nuclear waste. However, most studies exploring differences between white-males and minority populations have focused primarily on African American, Hispanic, and Asian populations with little understanding of how Native American groups perceive risk relative to other stakeholders. Trust in government and past experiences with environmental hazards may play an especially influential role in shaping the collective perceptions of risk among Native American populations who have been an important stakeholder in meeting the energy and resource needs of the United States. 


## Native American Risk Perceptions and the WIPP Facility 

Native Americans have a unique relationship with the federal government and have been at the heart of many environmental justice disputes over the past several decades. Therefore it is vitally important that we understand this particular group’s preferences on issues concerning nuclear waste. Based on previous research and important cultural and historical experiences with environmental justice issues, we expect that Native Americans will have risk perceptions closer to those of other minorities. The Native American population within New Mexico consists of 22 tribes and Pueblos including the Navajo Nation and the Mescalero Apache. Both the Navajo and Mescalero Apache have had difficult experiences with nuclear technology in the past, making it more likely that Native Americans in New Mexico will not be supportive of a nuclear waste facility. 

> H1: Native Americans will be less supportive and have higher risk perceptions of the Waste Isolation Pilot Plant (WIPP) facility than Whites.

<!--Furthermore, based on experiences with environmental justice issues that set Native Americans apart from other populations we expect risk perceptions to be even higher than that of other minority groups.
H2: Native Americans will be less supportive and have higher risk perceptions of WIPP than other minorities. -->

Given the dearth of quantitative analysis of the risk perceptions of Native populations, our analysis of intraracial differences among Native Americans is largely exploratory. However, based on the findings of @rivers_simple_2010
regarding African-Americans and gender, there are likely to be differences within the Native American sub-sample. 

> H2: Native American males will be more supportive and view less risk associated with WIPP than Native American women.

In addition, given the past experiences of the Navajo tribe with uranium mining, those respondents living in Navajo territory are likely to have different views about WIPP. 

> H3: Native Americans living in counties that contain Navajo land will be less supportive and perceive higher risk related to WIPP than Native Americans in other parts of New Mexico.

Finally, when examining the perceptions of Native Americans toward WIPP we will use several of the standard measures commonly used to predict risk perceptions including other demographics (age, education, income), values (political ideology, partisanship, environmental concern) and views of government. The next section introduces the data and methods used to explore these patterns among Native American populations. 

# Data and Methods 

The data that will be used comes from a collection of 35 statewide
surveys conducted in New Mexico, between 1990 and 2001, by the
University of New Mexico's Institute for Public Policy (IPP). These
surveys were collected quarterly until 1995 and were subsequently
administered twice a year. For the purposes of analysis, the surveys
are aggregated by year. During this period, data was collected from
over $\sim25,000$ respondents, including $\sim850$ that
self-identified as American Indian. The large sample of Native Americans allows for statistical analysis of this population.

We examine two dependent variables related to the WIPP facility. The first is a measure of support for WIPP and the second is a risk measure that measures the perceived risks associated with the transportation of nuclear material to the WIPP facility. WIPP support is based on the following question:

> There has been much controversy over WIPP, the Waste Isolation Plant in New Mexico, which is to serve as a disposal facility for transuranic radioactive wastes. Which of the following best represents your view? 

\noindent  Respondents were given four choices; open, open with minor change, open with major changes, or not open at all.^[For the surveys administered after WIPP opened, wording was changed to _should never have been opened_.] The responses were re-coded into a dichotomous variable so that one is WIPP should open/open with minor changes and zero is WIPP needs major changes or should not open. The mean for WIPP support is `r round(mean(wipp$wipp1.dic,na.rm=T),3)`, which indicates that when pooling the sample across all years `r round(mean(wipp$wipp1.dic,na.rm=T),3)*100`% of respondents were supportive of the WIPP facility. 

The transportation risk question asked respondents to rate the risks associated with the transportation of radioactive material to the WIPP facility:

> Thinking specifically about the transportation of wastes to WIPP, some opponents of WIPP have argued that the transportation of material to WIPP poses a significant risk of releasing radiation into the environment.

> The Department of Energy has argued that these risks are extremely small. Using a scale where one is no risk, five is extreme risk, and you may choose any number from one to five, how risky do you consider the transportation of low-and medium-level radioactive material to the WIPP facility to be? 

\noindent The mean for the transportation risk variable is `r round(mean(wipp$wipp2,na.rm=T),3)` on the one to five scale. The WIPP support question was asked in all years from 1990 to 2001 and the risk of transportation question was asked beginning in 1992. 

<!--^[Full question wording
is available in the appendix.]--> 

The main independent variables of interest are the racial demographics
of the respondents. To test the first hypothesis (H1) a dummy variable was created for Native Americans, with those that self-identified as American Indian coded as one. The mean of the Native American variable was `r round(mean(wipp$native,na.rm=T),3)`, indicating that `r round(mean(wipp$native,na.rm=T),3)*100`% of the sample identified as American Indian. In addition, a dummy variable for minority respondents, which includes self-identified African--Americans, Hispanics, and Asians, was created, and self-identified White respondents is used as the excluded referent category.

We also control for other demographic variables, such as
gender (Male = 1), education (a one to six scale, with one indicating some high school to six indicating post-graduate degree), and income (a one to six scale, in $10,000 increments). In addition, we include measures of values including political ideology--a seven point scale with one indicating strong conservative and seven indicating strong liberal--political partisanship--a dummy variable for the Republican party with one indicating the respondent self-identifies as a Republican and zero indicating Democratic party or other party. We also include a measure of environmental concern, with those that identified environmental quality as the most important problem facing New Mexico coded as one and those that indicated other issues as zero. As a proxy for trust in government, we use a measure of the performance of the New Mexico government. This measure is an average of three questions that asked respondents to rate, from one (poor) to four (excellent), the performance of New Mexico's legislature, public employees, and governor.^[Descriptive statistics for all variables are in the Appendix.] Finally, we control for year that the survey was administered.^[We use fixed effects for year, with a dummy variable for each year and with 1990 as the excluded referent year.]

# Results

As noted, it is expected that Native Americans would be less supportive of WIPP than Whites, all else equal (H1). To test this hypothesis, Logit regression was used, and the results are presented in Table \ref{tab:wipp1}. 

__Table \ref{tab:wipp1} about here__ 

As expected, we find that Native American respondents expressed less support for WIPP than did white respondents. Also and consistent with other research, other minority respondents were less supportive of the WIPP facility than whites. In addition, older, male, and higher income respondents were supportive of WIPP. With respect to values, we find that more conservative and Republican respondents were also more supportive than those with other political values, whereas those that were concerned about environmental quality were less supportive then those that rated another issue as a major concern. Finally, we find that opinions about the performance of the New Mexico state government was a positive and significant predictor of WIPP support. 

Next, we examine the perceived level of risk associated with the transportation of transuranic waste to the WIPP facility using OLS regression. 
The OLS model results are shown in Table \ref{tab:wipp2}.

__Table \ref{tab:wipp2} about here__ 

As expected, Native Americans, and other racial minorities rated the risks of transportation as higher, on average, than white respondents. Also as expected, older, male, higher educated, higher income respondents were less likely than other demographic groups to perceive risk associated with transportation to the WIPP facility. As respondents became more conservative, Republican, and had a positive view of New Mexico government performance, they were, on average, less likely to view transportation as risky. Finally, those concerned with environmental quality were more likely to see the transport of materials to WIPP as risky. 

To illustrate the results, we ran 10,000 simulations of the WIPP support and
transportation risk models. The simulations were used to estimate the model parameters [@king_making_2000; @gelman_data_2006].^[Based on the point estimates and the variance-covariance estimates from the models, the simulations take 10,000 random number draws to estimate the parameters.] The parameter estimates were then used to estimate the predicted probability of support for WIPP and the level of risk, from one to five, associated with the transportation of material to WIPP. The model simulations allow for the illustration of model uncertainty relative to the models predictions. Models were estimated for Native Americans, White males, and other minorities. The other variables were held constant at their mean, and the year variable was held constant at its median (1996). The models are shown in Figure \ref{fig:sims}.


```{r, include=FALSE}
source("nativeSIMScode.R")
pdf("supsims.pdf")
plot(density(white.sim1.P), xlim=c(0.2,0.8), ylim=c(0,40), main="", xlab="Density Curves of the Predicted Probabilities of WIPP Support")
lines(density(native.sim1.P), lty=2)
lines(density(minority.sim1.P), lty=3)
legend("topleft", c("White Males","Native Americans","Other Minorities"), lty=1:3, bty="n")
dev.off()
```

```{r, include=FALSE}
pdf("risksims.pdf")
plot(density(white.sim2), xlim=c(1,5), ylim=c(0,15), xaxt="n", main="", xlab="Density Curves of Perceived Risk of Transportation to WIPP")
axis(side = 1, at=1:5, label=c("No Risk","2","3","4","Extreme Risk"))
lines(density(native.sim2), lty=2)
lines(density(minority.sim2), lty=3)
legend("topleft", c("White Males","Native Americans","Other Minorities"), lty=1:3, bty="n")
dev.off()
```

__Figure \ref{fig:sims} about here__ 


As can be seen in Figure \ref{fig:supsims}, the predicted probability of support for WIPP of White males exceeds that of Native Americans and other minorities. The mean predicted probability for White males was `r round(mean(white.sim1.P),3)` with a range from `r round(min(white.sim1.P),3)` to `r round(max(white.sim1.P),3)`. For Native Americans the mean was `r round(mean(native.sim1.P),3)` (`r round(range(native.sim1.P),3)`) and for other minorities it was `r round(mean(minority.sim1.P),3)` (`r round(range(minority.sim1.P),3)`). Also as shown, support for WIPP between Native Americans and other minorities overlaps, indicating that no significant differences in support exist between those groups. In addition, as illustrated in Figure \ref{fig:risksims}, white males
tended to perceive the risks associated with transportation as less, at a mean of `r round(mean(white.sim2),3)` (`r round(range(white.sim2),3)`), than Native American and minority respondents with means of `r round(mean(native.sim2),3)` (`r round(range(native.sim2),3)`) and `r round(mean(minority.sim2),3)` (`r round(range(minority.sim2),3)`) respectively. As show, there is again overlap between Native Americans and other minorities concerning transportation risk, indicating no difference in risk perceptions between those two groups. 

## Native Americans 

```{r, include=FALSE}
library(car)
wipp$native.nw <- recode(wipp$county,                                "16=1;18=1;22=1;24=1;25=1;else=0")
```

While a large body of research has shown that white males tend to be more accepting of risk than racial minorities, less work has examined differences concerning risk perceptions _within_ minority groups, particularly Native Americans. To examine differences in risk perceptions between Native Americans, we use only those respondents that self-identified as American Indian. Building on @rivers_simple_2010, we expect Native American males to be more supportive of WIPP than Native American females (H2). In addition, we expect differences among tribes, particularly the Navajo given their history of adverse effects resulting from uranium mining (H3). As a proxy for Navajo tribal affiliation, we code the five counties (Los Alamos, McKinley, Rio Arriba, Sandoval, and San Juan) in the northwest corner of the state that overlap Navajo territory as one and all other counties a zero. The mean for the northwest county variable is `r round(mean(wipp$native.nw,na.rm=T),3)`, which indicates that `r round(mean(wipp$native.nw,na.rm=T),3)*100`% of Native American respondents live in those counties. As before, we first examine support for WIPP, and the results are shown in Table \ref{tab:native1}.

__Table \ref{tab:native1} about here__ 

As expected, we find that Native Americans males were, on average, more supportive of WIPP than Native American females. In addition, older Native Americans, those with higher incomes, and Republicans were also more likely to be supportive of WIPP. However, there were no significant differences between Native Americans in the northwest New Mexico counties and those in the rest of the state. Next we examine the risk perceptions associated with the
transportation of radioactive materials, and the results are shown in Table
\ref{tab:native2}.

__Table \ref{tab:native2} about here__

As can be seen, older Native American males with higher incomes tended to view the risks associated with transportation as less. In addition, Native Americans concerned about the environment were likely, at the _p_$<.10$ level, to see more risk from transportation. Finally, Native Americans that live
in the northwest counties tended to view the risks associated with
transportation as higher than those in other parts of the state. The finding that those in the Navajo portion of the state had differing views of risk points to intratribal differences concerning risk perceptions. However, more work is needed to explore these possible differences. In addition, the northwest counties are the farthest from the facility, therefore that finding lends support to the notion that, in contrast to the standard NIMBY model, those in closer proximity are more likely to be supportive and perceive less risk than those farther away. 

# Discussion and Conclusion 

The views of Native Americans regarding the risks associated with energy and environmental policy are particularly important to consider given the unique position of tribes within the intergovernmental framework of the US and the history of environmental justice concerns that continue to this day. Despite these concerns, little quantitative analysis of the risk perceptions of Native Americans has been done. Using environmental justice concerns related to Native American populations and the extant literature on risk perceptions, we derived some expectations concerning the perceptions of risk of the Native American population and, more specifically, options about the WIPP facility, a transuranic waste facility located in southeast New Mexico. It was posited that, because of their experiences and the white-male effect, Native Americans would be less supportive of WIPP and perceive more risk associated with WIPP than whites. This expectation was met as Native American were, _ceteris paribus_, significantly less supportive of WIPP than whites and significantly more likely to view the transportation of nuclear materials to the WIPP facility as posing more risk than whites. However, as evidenced by Figure \ref{fig:sims}, no significant differences existed between Natives and other minorities in their views of the WIPP facility. The lack of difference between Native Americans and other minorities could be seen as evidence of the role of power and status in society as a partial driver of the white-male effect. While Natives Americans and other racial minorities have differing experiences, both groups have historically been denied access to public decision-making processes and this historic lack of access can foster distrust of government activities and exacerbate perceptions of risk. In addition, white-males typically have higher incomes, more education, and are more politically conservative than other groups and these factors also contribute to the support of nuclear technology. 

<!-- While the major interest of this paper was the risk perceptions of Native Americans, we also controlled for other variables that have been found, in the extant literature, to influence risk perceptions. Specifically, we controlled for political values, environmental concern, and views about the performance of the New Mexico state government. In each case we found these to be significant predictors of views about WIPP. Specifically we found that increased conservatism, Republican party affiliation, and increased support of New Mexico government to be related to support for WIPP and less perception of risk related to transportation to WIPP. Whereas, we found that increased concern for the environment, on average, led to less support of WIPP and higher perceptions of transportation risk. -->

Apart from comparing Native Americans to other racial groups, we also examined differences within the Native American sub-sample. To our knowledge, no other study has used large-_N_ statistics to examine risk perceptions of Native Americans. Drawing on work examining differences in African-Americans [@rivers_simple_2010], we posited that Native American males, like African-Americans males in the previous study, would differ in their views of WIPP from Native American females, and this expectation was met. Specifically, we found that Native American males were more supportive and viewed less risk associated with WIPP than Native American females. In addition, we found that older and higher income Native Americans were more supportive and perceived less risk than those that were younger and lower income. With regard to values, Native Americans that identified as Republican were more supportive of WIPP and those that were more concerned with the environment were more likely (at $p<.10$) to view higher risks associated with transportation. Perhaps the most interesting observation concerned the existence of possible intertribal differences between groups and regions in New Mexico. We posited that Native Americans living in counties in the northwest part of New Mexico that contain Navajo land would be less supportive of WIPP than those in other parts of the state due to an extensive history of environmental issues in these areas, particularly the history of uranium mining in the Navajo territories. While we found no differences between regions on opinions toward WIPP in particular, we did find significantly higher perceptions of risk associated with the transportation of nuclear waste among Native populations in northwestern New Mexico. This would suggest that the collective experience of Native people toward issues of environmental risk are not uniform and may be shaped by more local historical events. 

There are several avenues for future research that deserve additional attention given limitations of the present study. First, a greater understanding of cultural differences across Native groups on risk is warranted in light of our findings concerning geographic variation in New Mexico. Admittedly, the county measure is a rough proxy for tribal affiliation. Based on the nature of the current dataset, it is not possible to place individual Native respondents to a particular nation, therefore, more work is needed to determine differences in risk perceptions across tribes using new and innovative survey instruments. Future research could also benefit from a more nuanced understanding of risk perception among Native populations across a wide range of environmental issues. The present investigation focuses specifically on nuclear waste and the WIPP facility, but as mentioned earlier Native people have struggled with a wide range of devastating environmental issues outside of this particular area. How these experiences shape Native perceptions of environmental risk is a much-needed area of future research. With these limitations noted, the present study greatly expands our understanding of risk among American Indian populations in America. 

\newpage

```{r, include=FALSE}
wippSUP <- glm(wipp1.dic ~
           age+gender+educ+inc50+minority+native+
           ideology+rep+env_bigprob+avg_gov_trust+as.factor(year),
           data=wipp, family=binomial())
library("apsrtable")
apsrtable(wippSUP, stars="default", digits=3, model.names=c("WIPP Support"), coef.names=c("Intercept","Age","Male","Education","Income","Minority","Native American","Ideology","Republican", "Environmental Concern","NM Government Performance"), omitcoef=list(12:21), caption=c("Logit Estimates of Support for the WIPP Facility"), label="tab:wipp1", notes=list("Estimated coefficients for survey years not listed", se.note, stars.note))
```

\begin{table}[!h] \centering \footnotesize
\caption{Logit Estimates of Support for the WIPP Facility}
\label{tab:wipp1} 
\begin{tabular}{ l D{.}{.}{3} } 
\hline 
  & \multicolumn{ 1 }{ c }{ WIPP Support } \\ \hline
 %                         & WIPP Support \\ 
Intercept                 & -2.898 ^{***}\\ 
                          & (0.135)      \\ 
Age                       & 0.015 ^{***} \\ 
                          & (0.001)      \\ 
Male                      & 0.825 ^{***} \\ 
                          & (0.035)      \\ 
Education                 & 0.014        \\ 
                          & (0.013)      \\ 
Income                    & 0.131 ^{***} \\ 
                          & (0.011)      \\ 
Minority                  & -0.466 ^{***}\\ 
                          & (0.039)      \\ 
Native American           & -0.636 ^{***}\\ 
                          & (0.097)      \\ 
Ideology                  & 0.118 ^{***} \\ 
                          & (0.012)      \\ 
Republican                & 0.446 ^{***} \\ 
                          & (0.039)      \\ 
Environmental Concern     & -0.722 ^{***}\\ 
                          & (0.118)      \\ 
NM Government Performance & 0.257 ^{***} \\ 
                          & (0.032)       \\
 $N$                       & 15818        \\ 
AIC                       & 19620.799    \\ 
BIC                       & 20264.987    \\ 
$\log L$                 & -9726.399     \\ \hline
 \multicolumn{2}{l}{\tiny{Estimated coefficients for survey years not listed}}\\
\multicolumn{2}{l}{\tiny{Standard errors in parentheses}}\\
\multicolumn{2}{l}{\tiny{$^\dagger$ significant at $p<.10$; $^* p<.05$; $^{**} p<.01$; $^{***} p<.001$}} 
\end{tabular} 
 \end{table}

```{r, include=FALSE}
wippRSK <- lm(wipp2 ~
          age+gender+educ+inc50+minority+native+
          ideology+rep+env_bigprob+avg_gov_trust+as.factor(year), data=wipp)
apsrtable(wippRSK, stars="default", digits=3, model.names=c("Transportation Risk"), coef.names=c("Intercept","Age","Male","Education","Income","Minority","Native American","Ideology","Republican", "Environmental Concern", "NM Government Performance"), omitcoef=list(12:21), caption=c("OLS Estimates of Perceived Risk of Nuclear Waste Transportation to the WIPP Facility"), label="tab:wipp2", notes=list("Estimated coefficients for survey years not listed", se.note, stars.note))
```

\newpage
\begin{table}[!h] \centering \footnotesize
\caption{OLS Estimates of Perceived Risk of Nuclear Waste Transportation to the WIPP Facility}
\label{tab:wipp2} 
\begin{tabular}{ l D{.}{.}{3} } 
\hline 
  & \multicolumn{ 1 }{ c }{ Transportation Risk } \\ \hline
 %                         & Transportation Risk\\ 
Intercept                 & 4.830 ^{***}       \\ 
                          & (0.079)            \\ 
Age                       & -0.008 ^{***}      \\ 
                          & (0.001)            \\ 
Male                      & -0.596 ^{***}      \\ 
                          & (0.021)            \\ 
Education                 & -0.036 ^{***}      \\ 
                          & (0.008)            \\ 
Income                    & -0.085 ^{***}      \\ 
                          & (0.007)            \\ 
Minority                  & 0.365 ^{***}       \\ 
                          & (0.024)            \\ 
Native American           & 0.500 ^{***}       \\ 
                          & (0.058)            \\ 
Ideology                  & -0.062 ^{***}      \\ 
                          & (0.007)            \\ 
Republican                & -0.280 ^{***}      \\ 
                          & (0.024)            \\ 
Environmental Concern     & 0.443 ^{***}       \\ 
                          & (0.072)            \\ 
NM Government Performance & -0.146 ^{***}      \\ 
                          & (0.020)             \\
 $N$                       & 13409              \\ 
$R^2$                     & 0.174              \\ 
adj. $R^2$                & 0.173              \\ 
Resid. sd                 & 1.212               \\ \hline
 \multicolumn{2}{l}{\tiny{Estimated coefficients for survey years not listed}}\\
\multicolumn{2}{l}{\tiny{Standard errors in parentheses}}\\
\multicolumn{2}{l}{\tiny{$^\dagger$ significant at $p<.10$; $^* p<.05$; $^{**} p<.01$; $^{***} p<.001$}} 
\end{tabular} 
 \end{table}

```{r, include=FALSE}
wipp1.native <- glm(wipp1.dic ~ age+gender+educ+inc50+ideology+rep +env_bigprob+avg_gov_trust+native.nw+as.factor(year),data=subset(wipp, native==1), family=binomial())
library(apsrtable)
apsrtable(wipp1.native, stars="default", digits=3, model.names=c("WIPP Support"), coef.names=c("Intercept","Age","Male","Education","Income","Ideology","Republican", "Environmental Concern", "NM Government Performance", "Northwest County"), omitcoef=list(11:20), caption=c("Logit Estimates of Support for the WIPP Facility: Native Americans"), label="tab:native1", notes=list("Estimated coefficients for survey years not listed", se.note, stars.note))
```

\newpage
\begin{table}[!h] \centering \footnotesize
\caption{Logit Estimates of Support for the WIPP Facility: Native Americans}
\label{tab:native1} 
\begin{tabular}{ l D{.}{.}{3} } 
\hline 
  & \multicolumn{ 1 }{ c }{ WIPP Support } \\ \hline
 %                         & WIPP Support \\ 
Intercept                 & -2.337 ^{***}\\ 
                          & (0.693)      \\ 
Age                       & 0.021 ^{**}  \\ 
                          & (0.007)      \\ 
Male                      & 0.441 ^*     \\ 
                          & (0.193)      \\ 
Education                 & -0.049       \\ 
                          & (0.071)      \\ 
Income                    & 0.164 ^{**}  \\ 
                          & (0.063)      \\ 
Ideology                  & 0.054        \\ 
                          & (0.061)      \\ 
Republican                & 0.549 ^*     \\ 
                          & (0.223)      \\ 
Environmental Concern     & -0.142       \\ 
                          & (0.578)      \\ 
NM Government Performance & 0.055        \\ 
                          & (0.181)      \\ 
Northwest County          & -0.247       \\ 
                          & (0.201)       \\
 $N$                       & 566          \\ 
AIC                       & 702.223      \\ 
BIC                       & 1049.311     \\ 
$\log L$                 & -271.112      \\ \hline
 \multicolumn{2}{l}{\tiny{Estimated coefficients for survey years not listed}}\\
\multicolumn{2}{l}{\tiny{Standard errors in parentheses}}\\
\multicolumn{2}{l}{\tiny{$^\dagger$ significant at $p<.10$; $^* p<.05$; $^{**} p<.01$; $^{***} p<.001$}} 
\end{tabular} 
 \end{table}

```{r, include=FALSE}
wipp2.native <- lm(wipp2 ~ age+gender+educ+inc50+ideology+rep+env_bigprob +avg_gov_trust+native.nw+as.factor(year),data=subset(wipp, native==1))
apsrtable(wipp2.native, stars="default", digits=3, model.names=c("Transportation Risk"), coef.names=c("Intercept","Age","Male","Education","Income","Ideology","Republican", "Environmental Concern","NM Government Performance","Northwest County"), omitcoef=list(11:20), caption=c("OLS Estimates of Perceived Risk of Nuclear Waste Transportation to the WIPP Facility: Native Americans"), label="tab:native2", notes=list("Estimated coefficients for survey years not listed", se.note, stars.note))
```

\newpage
\begin{table}[!h] \centering \footnotesize
\caption{OLS Estimates of Perceived Risk of Nuclear Waste Transportation to the WIPP Facility: Native Americans}
\label{tab:native2} 
\begin{tabular}{ l D{.}{.}{3} } 
\hline 
  & \multicolumn{ 1 }{ c }{ Transportation Risk } \\ \hline
 %                         & Transportation Risk\\ 
Intercept                 & 4.974 ^{***}       \\ 
                          & (0.442)            \\ 
Age                       & -0.020 ^{***}      \\ 
                          & (0.004)            \\ 
Male                      & -0.305 ^*          \\ 
                          & (0.127)            \\ 
Education                 & 0.029              \\ 
                          & (0.046)            \\ 
Income                    & -0.092 ^*          \\ 
                          & (0.040)            \\ 
Ideology                  & 0.013              \\ 
                          & (0.040)            \\ 
Republican                & -0.254             \\ 
                          & (0.154)            \\ 
Environmental Concern     & 0.745 ^\dagger    \\ 
                          & (0.395)            \\ 
NM Government Performance & -0.108             \\ 
                          & (0.119)            \\ 
Northwest County          & 0.389 ^{**}        \\ 
                          & (0.129)             \\
 $N$                       & 468                \\ 
$R^2$                     & 0.141              \\ 
adj. $R^2$                & 0.109              \\ 
Resid. sd                 & 1.315               \\ \hline
 \multicolumn{2}{l}{\tiny{Estimated coefficients for survey years not listed}}\\
\multicolumn{2}{l}{\tiny{Standard errors in parentheses}}\\
\multicolumn{2}{l}{\tiny{$^\dagger$ significant at $p<.10$; $^* p<.05$; $^{**} p<.01$; $^{***} p<.001$}} 
\end{tabular} 
 \end{table}

\newpage
\begin{figure}[!h]
\centering
\caption{Simulations of Support for WIPP and Transportation Risk} \label{fig:sims}
\begin{subfigure}[b]{0.45\textwidth}
\caption{WIPP Support}
\label{fig:supsims}
\includegraphics[width=\textwidth]{supsims.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
\caption{Transportation Risk}
\label{fig:risksims}
\includegraphics[width=\textwidth]{risksims.pdf}
\end{subfigure}
\end{figure}

\newpage

# Appendix {.unnumbered}
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}

<!-- descriptive stats -->
```{r, include=FALSE}
install.packages("stargazer", dependencies=TRUE, repos='http://cran.us.r-project.org')
library(stargazer)
vars.dat <- data.frame(wipp$age, wipp$gender, wipp$educ, wipp$inc50, wipp$minority, wipp$native, wipp$ideology, wipp$rep, wipp$avg_gov_trust, wipp$wipp1.dic, wipp$wipp2)
stargazer(vars.dat, covariate.labels=c("Age","Male","Education","Income","Minority","Native American","Ideology","Republican", "NM Government Performance", "WIPP Support", "Transportation Risk"), title="Descriptive Statistics for Entire Sample", label="tab:descr")
```

\begin{table}[!htbp] \centering 
  \caption{Descriptive Statistics} 
  \label{tab:descr} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Age & 23,629 & 44.645 & 16.651 & 18 & 99 \\ 
Male & 23,870 & 0.430 & 0.495 & 0 & 1 \\ 
Education & 23,724 & 3.678 & 1.524 & 1 & 6 \\ 
Income & 20,821 & 3.819 & 1.678 & 1 & 6 \\ 
Minority & 22,702 & 0.330 & 0.470 & 0 & 1 \\ 
Native American & 22,702 & 0.036 & 0.186 & 0 & 1 \\ 
Ideology & 22,280 & 4.276 & 1.571 & 1 & 7 \\ 
Republican & 21,416 & 0.355 & 0.479 & 0 & 1 \\ 
NM Government Performance & 20,677 & 2.202 & 0.550 & 1.000 & 4.000 \\ 
WIPP Support & 21,543 & 0.478 & 0.500 & 0 & 1 \\ 
Transportation Risk & 17,958 & 3.138 & 1.341 & 1 & 5 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 

```{r, include=FALSE}
library(stargazer)
vars.dat2 <- data.frame(wipp$age, wipp$gender, wipp$educ, wipp$inc50, wipp$ideology, wipp$rep, wipp$avg_gov_trust, wipp$wipp1.dic, wipp$wipp2)
vars.dat2 <- subset(vars.dat2, wipp$native==1)
stargazer(vars.dat2, covariate.labels=c("Age","Male","Education","Income","Ideology","Republican", "NM Government Performance", "WIPP Support", "Transportation Risk"), title="Descriptive Statistics: Native Americans", label="tab:descr2")
```

\begin{table}[!htbp] \centering 
  \caption{Descriptive Statistics: Native Americans} 
  \label{tab:descr2} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Age & 808 & 38.066 & 14.512 & 18 & 85 \\ 
Male & 815 & 0.421 & 0.494 & 0 & 1 \\ 
Education & 811 & 3.298 & 1.476 & 1 & 6 \\ 
Income & 747 & 3.264 & 1.674 & 1 & 6 \\ 
Ideology & 781 & 4.193 & 1.586 & 1 & 7 \\ 
Republican & 729 & 0.240 & 0.427 & 0 & 1 \\ 
NM Government Performance & 722 & 2.224 & 0.546 & 1.000 & 4.000 \\ 
WIPP Support & 773 & 0.336 & 0.473 & 0 & 1 \\ 
Transportation Risk & 634 & 3.609 & 1.372 & 1 & 5 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 


\newpage

# References {.unnumbered}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\singlespace
\noindent



